Deep Reinforcement Learning for Conservation Decisions

### 2021-06-11 manuscript/manuscript.Rmd

---
title: Deep Reinforcement Learning for Conservation Decisions
authors:
  - name: Marcus Lapeyrolerie
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
  - name: Melissa Chapman
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
  - name: Kari Norman
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
  - name: Carl Boettiger
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: cboettig@berkeley.edu
abstract: |
  Enter the text of your abstract here.
keywords:
  - blah
  - blee
  - bloo
  - these are optional and can be removed
bibliography: references.bib
nocite: |
    @A2C, @trpo, @ddpg, @TD3, @impala
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
output: 
  rticles::arxiv_article:
    keep_tex: true
  word_document: default
editor_options: 
  markdown: 
    wrap: sentence
---

@@

Our second example focuses on a case for which we do not have an existing, provably optimal policy to compare against.
We consider the generic problem of an ecosystem facing slowly deteriorating environmental conditions which move the dynamics closer towards a tipping point [Fig 2].
This model of a critical transition has been posited widely in ecological systems, from the simple consumer-resource model of [@May1977] on which our dynamics are based, to microbial dynamics [@Dai2012], lake ecosystem communities [@Carpenter2011] to planetary ecosystems [@Barnosky2012].
On top of these ecological dynamics we introduce an explicit ecosystem service model quantifying the value of more desirable 'high' state relative to the 'low' state.
For simplicity, we assume a proportional benefit $b$ associated with the ecosystem state $X(t)$.
Thus when the ecosystem is near the 'high' equilibrium $\hat X_H$, the corresponding ecosystem benefit $b \hat X_H$ is higher than at the low equilibrium, $b x_L$, consistent with the intuitive description of ecosystem tipping points [@Barnosky2012].

@@

Examining the replicate management trajectories and corresponding rewards [Fig 3] reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping the cumulative reward into negative territory on average and well below the reward realized by the simple steady-state strategy. <!-- ML: Nit picky; it looks like it almost goes negative but not quite on average in the plot -->
This initial investment then begins to pay off -- by about the 200th time step the RL agent has surpassed the performance of the steady state strategy.
The policy plot provides more intuition for the RL agent's strategy: at very high state values, the RL agent opts for no conservation action -- so far from the tipping point, no response is required.
Near the tipping point, the RL agent steeply ramps up the conservation effort, and retains this effort even as the system falls below the critical threshold, where a sufficiently aggressive response can tip the system back into recovery.
For a system at or very close to the zero-state, the RL agent gives up, opting for no action.
Recall that the quadratic scaling of cost makes the rapid response of the TD3 agent much more costly to achieve the same net environmental improvement divided into smaller increments over a longer timeline.
However, our RL agent has discovered that the extra investment for a rapid response is well justified as the risk of crossing a tipping point increases.

A close examination of the trajectories of individual simulations which cross the tipping point under either management strategy [see appendix B] further highlights the difference between these two approaches.
Under the steady-state strategy, the system remains poised too close to the tipping point: stochastic noise eventually drives most replicates across the threshold, where the steady-state strategy is too weak to bring them back once they collapse.
As replicate after replicate stochastically crashes, the mean state and mean reward bend increasingly downwards.
In contrast, the RL agent edges the system slightly farther away from the tipping point, decreasing but not eliminating the chance of a chance transition.
In the few replicates that experience a critical transition anyway, the RL agent usually responds with sufficient commitment to ensure their recovery [Appendix B].
Only 3 out of 100 replicates degrade far enough for the RL agent to give up the high cost of trying to rescue them.
The RL agent's use of a more dynamic strategy out-preforms the steady-state strategy.
Numerous kinks visible in the RL policy function also suggest that this solution is not yet optimal.
Further tuning of hyper-parameters, increased training, alterations or alternatives to the training algorithm would likely be able to further improve upon this performance.
Nevertheless, such quirks are likely to be common features of RL solutions -- long as they have minimal impact on realized rewards.

>>>>

---
title: Deep Reinforcement Learning for Conservation Decisions
authors:
  - name: Marcus Lapeyrolerie
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
  - name: Melissa S. Chapman
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
  - name: Kari E. A. Norman
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
  - name: Carl Boettiger
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: cboettig@berkeley.edu
abstract: |
  Can machine learning help us make better decisions about a changing planet?
  In this paper, we illustrate and discuss the potential of a promising 
  corner of machine learning known as _reinforcement learning_ (RL) to help tackle the
  most challenging conservation decision problems. RL is uniquely
  well suited to conservation and global change challenges for three reasons: (1)
  RL explicitly focuses on designing an agent who _interacts_ with an environment which is dynamic and uncertain,
  (2) RL approaches do not require massive amounts of data, (3) RL approaches would 
  utilize rather than replace existing models, simulations, and the knowledge they contain.
  We provide a conceptual and technical introduction to RL and its relevance to ecological
  and conservation challenges, including examples of a problem in setting
  fisheries quotas and in managing ecological tipping points.
  Four appendices with annotated code provide a tangible introduction to researchers 
  looking to adopt, evaluate, or extend these approaches.

keywords:
  - Reinforcement Learning
  - Conservation
  - Machine Learning
  - Artificial Intelligence
  - Tipping points
bibliography: references.bib
nocite: |
    @A2C, @trpo, @ddpg, @TD3, @impala
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
output: 
  rticles::arxiv_article:
    keep_tex: true
  word_document: default
editor_options: 
  markdown: 
    wrap: sentence
---

@@

Our second example focuses on a case for which we do not have an existing, provably optimal policy to compare against.
We consider the generic problem of an ecosystem facing slowly deteriorating environmental conditions which move the dynamics closer towards a tipping point [Fig 4].
This model of a critical transition has been posited widely in ecological systems, from the simple consumer-resource model of [@May1977] on which our dynamics are based, to microbial dynamics [@Dai2012], lake ecosystem communities [@Carpenter2011] to planetary ecosystems [@Barnosky2012].
On top of these ecological dynamics we introduce an explicit ecosystem service model quantifying the value of more desirable 'high' state relative to the 'low' state.
For simplicity, we assume a proportional benefit $b$ associated with the ecosystem state $X(t)$.
Thus when the ecosystem is near the 'high' equilibrium $\hat X_H$, the corresponding ecosystem benefit $b \hat X_H$ is higher than at the low equilibrium, $b x_L$, consistent with the intuitive description of ecosystem tipping points [@Barnosky2012].

@@

Examining the replicate management trajectories and corresponding rewards [Fig 5] reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping the cumulative reward into negative territory on average and well below the reward realized by the simple steady-state strategy. <!-- ML: Nit picky; it looks like it almost goes negative but not quite on average in the plot -->
This initial investment then begins to pay off -- by about the 200th time step the RL agent has surpassed the performance of the steady state strategy.
The policy plot provides more intuition for the RL agent's strategy: at very high state values, the RL agent opts for no conservation action -- so far from the tipping point, no response is required.
Near the tipping point, the RL agent steeply ramps up the conservation effort, and retains this effort even as the system falls below the critical threshold, where a sufficiently aggressive response can tip the system back into recovery.
For a system at or very close to the zero-state, the RL agent gives up, opting for no action.
Recall that the quadratic scaling of cost makes the rapid response of the TD3 agent much more costly to achieve the same net environmental improvement divided into smaller increments over a longer timeline.
However, our RL agent has discovered that the extra investment for a rapid response is well justified as the risk of crossing a tipping point increases.

A close examination of the trajectories of individual simulations which cross the tipping point under either management strategy [see Appendix B] further highlights the difference between these two approaches.
Under the steady-state strategy, the system remains poised too close to the tipping point: stochastic noise eventually drives most replicates across the threshold, where the steady-state strategy is too weak to bring them back once they collapse.
As replicate after replicate stochastically crashes, the mean state and mean reward bend increasingly downwards.
In contrast, the RL agent edges the system slightly farther away from the tipping point, decreasing but not eliminating the chance of a chance transition.
In the few replicates that experience a critical transition anyway, the RL agent usually responds with sufficient commitment to ensure their recovery [Appendix B].
Only 3 out of 100 replicates degrade far enough for the RL agent to give up the high cost of trying to rescue them.
The RL agent's use of a more dynamic strategy out-preforms the steady-state strategy.
Numerous kinks visible in the RL policy function also suggest that this solution is not yet optimal.
Further tuning of hyper-parameters, increased training, alterations or alternatives to the training algorithm would likely be able to further improve upon this performance.
Nevertheless, such quirks are likely to be common features of RL solutions -- long as they have minimal impact on realized rewards.
